Supplementary information - The Functional Therapeutic Chemical Classification System

### 2013-10-14 document.tex
correction DRS

This section describes the building mechanism behind the FTC.
First, a list of categories describing the mode and mechanism of action of drugs is defined. 
Then in a second step the newly created categories are automatically populated with approved compounds. 
Finally, the FTC is evaluated and repositioning hypotheses can be generated.

@@

\section{Equivalent definitions}
FTC classes are generated as presented in the previous section. Up to this point, these categories 
are only tokens with a human readable label as well as an identifier. The next step is going to 
assign equivalent definitions to each FTC class. A reasoner can understand such definitions and will automatically 
classify the knowledge base accordingly; drugs will then be assigned to FTC categories and the taxonomic structure 
arises after this reasoning step. Equivalent definitions are written as OWL class expressions using the entities of 
the knowledge base (summarised at {{https://github.com/loopasam/ftc/wiki/Knowledge-Base}} and in section \ref{specs}).
There are two types of equivalences: The first one captures perturbation of regulatory 
biological processes (so called \emph{regulatory patterns}) and the second one handles the perturbed functions (\emph{functional patterns}).

@@

\subsection{Gene Ontology annotations}
The GO annotation program aims to provide high-quality GO annotations to 
proteins in UniProt \citep{Dimmer2012}. In the context of the FTC, such annotations are used 
to create axioms linking protein targets to molecular functions and biological processes. 
Each protein annotated with a function creates an axiom such 
as \emph{[protein] SubClassOf has-function some [molecular function]}. 
Each protein annotated to a biological process creates an axiom such 
as \emph{[protein] SubClassOf involved-in some [biological process]}. 
E.g. \emph{Prothrombin SubClassOf involved-in some positive regulation of blood coagulation}. 
Each protein can be involved in multiple processes and capable of realizing multiple functions;
the polypharmacology is captured at this level.

@@

\section{Semantic similarity}
The semantic similarity measure performed over the FTC is a derivative of the 
Jaccard index \citep{jaccard1912distribution} \citep{rogers1960computer} (see \cite{pesquita2009semantic} for review). 
It is probably best understood as an example: If we consider two classes A and B, the semantic similarity 
between these classes corresponds to the number of OWL superclasses (direct and indirect, obtained with a reasoner) that 
are shared by A and B (intersection) divided by the number of superclasses of A or B (union). The index ranges 
from 0 (totally different) to 1 (identical).

>>>>

This section describes the building mechanism behind the FTC.
First, a list of \emph{categories} describing the mode and mechanism of action of drugs is defined (the word \emph{category} is interchangeable with \emph{term}, \emph{class} or \emph{concept} in this manuscript). 
Then in a second step the newly created categories are automatically populated with approved compounds. 
Finally, the FTC is evaluated and repositioning hypotheses can be generated.

@@

\section{Equivalent definitions}
FTC classes are generated as presented in the previous section. Up to this point, these categories 
are only tokens with a human readable label as well as an identifier. The next step is going to 
assign equivalent definitions to each FTC class. An OWL reasoner can understand such definitions and will automatically 
classify the knowledge base accordingly, following standard description logics reasoning services (see wiki for introduction).
Drugs will then be assigned to FTC categories and the taxonomic structure 
arises after this reasoning step. Equivalent definitions are written as OWL class expressions using the entities of 
the knowledge base (summarised at {{https://github.com/loopasam/ftc/wiki/Knowledge-Base}} and in section \ref{specs}).
There are two types of equivalences: The first one captures perturbation of regulatory 
biological processes (so called \emph{regulatory patterns}) and the second one handles the perturbed functions (\emph{functional patterns}).

@@

\subsection{Gene Ontology annotations}
The GO annotation program aims to provide high-quality GO annotations to 
proteins in UniProt \citep{Dimmer2012}. In the context of the FTC, such annotations are used 
to create axioms linking protein targets to molecular functions and biological processes. 
Each protein annotated with a function creates an axiom such 
as \emph{[protein] SubClassOf has-function some [molecular function]}. 
Each protein annotated to a biological process creates an axiom such 
as \emph{[protein] SubClassOf involved-in some [biological process]}. 
E.g. \emph{Prothrombin SubClassOf involved-in some positive regulation of blood coagulation}. 
Each protein can be involved in multiple processes and capable of realising multiple functions;
some of the polypharmacology is captured at this level.

@@

\section{Semantic similarity}
The semantic similarity measure performed over the FTC is a derivative of the 
Jaccard index \citep{jaccard1912distribution} \citep{rogers1960computer} (see \cite{pesquita2009semantic} for review). 
It is probably best understood as an example: If we consider two classes A and B, the semantic similarity 
between these classes corresponds to the number of OWL superclasses (direct and indirect, obtained with a reasoner) that 
are shared by A and B (intersection) divided by the number of superclasses of A or B (union). The index ranges 
from 0 (totally different) to 1 (identical). A similar approach was successfully implemented by \citep{hoehndorf2011phenomenet}, for similarity computations over phenotypic traits.
